Stochastic Optimal Control of Single Neuron Spike Trains

### 2013-08-29 draft.tex
Commiting old multi-train figs before pruning down

% \documentclass[12pt]{iopart}
\documentclass{article}

@@

\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Theorem}[thm]
% \theoremstyle{definition}
\newtheorem{ex}{Example}[thm]
\newtheorem{defn}{Definition}[thm]

@@

\title{Stochastic Optimal Control of Single Neuron Spike Trains
\\
---
\vskip5pt
Draft
}
\author{Alexandre Iolov, Susanne Ditlevsen, Andr\'e Longtin,
		}		
% \address{Department of Mathematics and Statistics, University of Ottawa, Ottawa, Canada 
% \\ Department of Mathematical Sciences, University of Copenhagen, Copenhagen,
% Denmark }
% \ead{$<$\href{mailto:aiolo040@uottawa.ca}
% 		{aiolo040 at uottawa dot ca}$>$}

\date{\today}

@@

\abstract{
{\em Objective.} External control of exact spike timings in single neurons is
important for basic understanding of neuronal activity, and has
potential for improving neural prostheses and in medical applications
related to neuronal disorders, such as Parkinson's disease or epilepsy.
The goal of this paper is the design of optimal controlled
electrical stimulation patterns to a cell to achieve a target
spike train under physiological constraints not to damage tissue and
minimize possible side effects.
{\em Approach.} We pose an optimal control problem to precisely specify
the spike timing in a noisy leaky integrate-and-fire model of a
neuron. In particular, we allow for the noise to be of arbitrary intensity. The optimal control problem is solved using Dynamic Programming if the controller has access to the voltage (closed-loop control) and using a Maximum Principle for the transition density if the controller only has access to the spike timings (open loop
control).
{\em Main results.} We have developed a stochastic optimal control
scheme to obtain precise spike timings, which is not limited
by the intensity of the noise in the system. Given this scope, the
numerical demands are non-trivial, but simulations show that the
algorithms are potentially feasible. It however remains to see if
they can be successfully applied in a real-time experimental setting.
Furthermore, we show how wrong the control strategy can perform if we
assume that the noise is negligible.
{\em Significance.} More precise targeting and control of neural
activity to achieve optimal performance in a noisy environment, in
which most neurons are embedded, have great potential in health
care. The main contribution is the online feedback control of a noisy
system through
modulation of the input, taking into account physiological constraints
on the control, as well as incurred costs to diminish adverse effects.
}

\tableofcontents

MSC KEYS: 49N90 90C39 93E20 93C20 93C95

@@

Our objective, namely targeting exact spike times in single neurons, has been
considered mainly in the open-loop context, and in either absence of or for
small noise. In \cite{Ahmadian2011} they use the Spike Response Model,
\cite{Gerstner2002}, to control output target spike trains and implement their
scheme on pyramidal cells in mouse cortical slices. Their method is numerically
efficient and allows for the simultaneous control of many neurons. However the
efficiency of the numerics strongly relies on the assumption that the diffusion
noise, the value of $\b$ in \cref{eq:X_evolution_uo} below, is small enough, see
the discussion above and below equation 12 in \cite{Ahmadian2011}. They only
work with open-loop control. The difference between the objective in
\cite{Ahmadian2011} and what we consider is that they maximize the probability
of spiking at some given time $t^*$, whereas we minimize the mean squared
difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
Moehlis et al. \cite{Moehlis2006} work with a Phase Response Model to obtain
spikes at exact times, while keeping the root mean square of the input to a
minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
consider noise, though. In \cite{Nabi2013}, the methods are implemented on brain
slices of pyramidal neurons of rat hippocampus. [[[ALEX, Is it closed- or open
loop? Could you SHORTLY describe the difference of what they do and what we
do?]]] \{ALEX: They use Phase Response Curves, assuming that a spike coincides
with $\th=0$; no noise, Pontryagin Maximum Principle, solved numerically $=>$
open-loop. Short Description: Since they do not consider noise, it is
straight-forward to make the neuron spike when you want it, and in fact that can
be done in many ways. Thus their optimization reduces to finding the way to make
the neuron spike at a pre-specified time, using least energy (the integral of
$u^2$.\}. While the Phase Response Curve Model is a parsimonious and effective
way to describe a neuron's response to a stimulus, it is only valid in the
supra-threshold regime, when the unstimulated neuron is periodically spiking.
In \cite{Feng2003} they use a leaky integrate-and-fire model and also
use open-loop control, but to obtain exact spike timings, they choose an
objective of minimizing the variance of the membrane potential at the desired
spike time, while forcing the mean of the membrane potential at this time point
to equal the threshold. This provides exact solutions since it does not involve
first-passage times, but has the drawback of getting a non-negligible
probability of being way-off the objective.

Our objective of imposing a certain timing sequence of the spike train using an
externally applied control is obtained in both the closed- and
open-loop setting, and we specifically include the noise in the
calculations of the controls. We restrict the controlled input to stay
within pre-specified bounds, and also include a cost function to
minimize intervention.
Our main contributions are... (((what we have done which has not been
done before, numerical merits...)))
\{ALEX: Basically, we deal with a significant noise component. Everything
else follows, i.e. the objective only makes sense if the ISI is a random variable.
The PDEs (HJB and Max Principle) only occur because of the noise
component, o/w we would have ODEs. The numerical schemes are simple given the
PDEs, we just use the Crank-Nicholson method for the PDEs and a basic
gradient-descent for the optimization in the open-loop case. The numerics take a
non-trivial time to run - hard to see them coming down to real-time, i.e.
10-20 ms, \ldots but possible. We also explore a wide spectrum of parameter
values, in particular we do  not restrict our attention to the
autonomously-spiking, supra-threshold regime.\}

[[[Susanne, we should mention Whittle \cite{Whittle1996}, but here, or together
with other monographs on optimal control?]]] \{ALEX: Indeed we should mention
Whittle, in particular mention that he discusses analytically the problem in the
closed-loop case (only), but not the open-loop case. He does not provide
numerics in either case. I would refrain from citing other books on
stochastic optimal control, here. Books commonly cited are
\cite{Fleming1975,Krylov2008} and the chapter on HJB equation in Oksendal,
\cite{Oksendal2007}, some of them are cited in later sections\ldots\}

@@

\section{Problem Formulation}
A basic but useful model for the neural membrane potential evolution is the
noisy leaky-integrate-and-fire model:
\begin{equation}
\begin{gathered} 
dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
\\
X(0) = 0,
\\
X(\ts) = \xth \implies   
\begin{cases}
X(\ts^+) = 0 &  
\end{cases}
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}
Here $X$ represents the membrane electric potential which tries to revert to $0$
with a time-const of $\tc$, $dW$ is a Brownian motion increment scaled by $\b$
and $\Iext(t)$ is the deterministic external input to the cell. Having last
spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
resets to $0$ and we start all over again. We write $\ts^+$ in the formula for
the reset since we are working in continuous time, i.e.\ it is really the
voltage's limit from the right at $\ts$ that is 0. For our purposes, we will
always have that the threshold is set to one, $\xth = 1$.

Suppose that we have some control over the external current such that it can be
decomposed as
\begin{equation}
\Iext(t) = \m + \a(t)
\label{eq:current_mu_alpha}
\end{equation}
where $\m$ is an uncontrollable, but constant part of the external current and
$\a(t)$ is controllable, i.e.\ it can be chosen to achieve some goal. 
A natural goal is to attempt to control the spike time, $\ts$. 
That is we would like to ask how do we choose $\a(t)$ such that $\ts = \T$.
A natural optimal control objective to achieve this is the least squares solution
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
\label{eq:OC_LS_variance}   
\end{equation}
where expectation is taken with respect to the distribution of the trajectories
of $X_t$.

Often the control has certain constraints. The most common are simple box
constraints:
\begin{equation}
\a(t) \in [\amin, \amax] \quad \forall t
\label{eq:bound_constraints_alpha}
\end{equation}

% \subsection{Optimal Control Formulation}
In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
running cost based on the control. This regularizes the problem eliminating the
subtleties of singular-control situations and it might be argued appropriate to
avoid excessive control. WE SHOULD DISCCUSS THIS IN GREATER DETAIL, BUT PERHAPS
LATER. (ALEX:\cite{Ahmadian2011} state that they impose a similar control
penalty to avoid excessive charge building up on the cell.)

So we seek an optimal control, $\a^*$, that solves
\begin{align}
J[\a(\cdot)] =&  
\Exp\left[
\e \int_0^\ts  \a^2(s) \intd{s}
+  
(\ts - \T \big)^2 \right]  
\label{eq:OC_LS_variance_energy}   
\\ \notag
\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)]. 
\end{align}
where $\e$ measure how much weight we put on minimizing the energy expenditure.
If $\e = 0$, then we do not care at all about the expended
energy.

It will often be the case that we write $\a$ and mean either a function or a
value of that function at a particular time. And that this function could be
random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
of $X$, it will be at least measurable with respect to the filtration of $X$,
see \cite{Krylov2008}. We will try to make clear below when we are refering to
$\a$ as a stochastic process and when we are merely refering to its particular
value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
refers to the function, while $\a(s)$ refers to that function's value, possibly
random, at time $s$.

We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
control. In closed-loop control, the value of $X_t$ at time $t$ is observable
and can be used in determining the control. In open-loop control only the spike
times are observable. In the closed-loop context, we might better write the
control as $$\a = \a(x,t)$$ to indicate its dependence on $X_t$ and to express
that it will be updated based on the time-course of $X$. 
In the open-loop context, we would instead write $$\a = \a(t)$$ to indicate that
$\a(\cdot)$ is decided for all times at time 0. 
The techniques used in the two scenarios will be different. For the closed-loop
scenario we will use Dynamic Programing, \cite{Fleming1975}, in order to obtain
the optimal control, while for the open-loop scenario we will use a form of the
Maximum Principle applied on the transition density of the controlled process,
\cite{Ahmed1981,Borzi2012}.

Crucially, we assume that the model parameters, $\m,\tc,\b$ in
\cref{eq:X_evolution_uo} are known.

\subsection{Parameter Regimes}
Different  parameter regimes can be envisioned given \cref{eq:X_evolution_uo},
depending on whether the noise intensity, $\b$, is
relatively high or low, and whether the external, uncontrollable bias current,
$\m$, induces spikes in the absence of noise or not.

Spikes will occur in the absence of noise, if and only
if $$ \m > \frac 1\tc,$$ which is called the supra-threshold regime. When $\m
\leq 1/\tc$, the regime is called the sub-threshold regime.
In addition we will impose two values of $\b$, which we
will call high-noise and low-noise, respectively.
Example values for each parameter regime are given in \cref{tab:regimes}, and we
visualize a single path for each regime in \cref{fig:regime_path_examples}.
\begin{table}
\begin{tabular}{|l||{c}|{c}|}
\hline
\backslashbox{$\m$}{$\b$}
& $1.5$ & $0.3$ \\
\hline
$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\ 
\hline
$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
\hline
\end{tabular}
\caption{Regime labels and example values. Note that for the numerical
experiments below, we use $\tc = 0.5$}
\label{tab:regimes}
\end{table}
%\usepackage{graphics} is needed for \includegraphics
\begin{figure}[htp]
\begin{center} 
  \includegraphics[width=0.99\textwidth]{Figs/PathSimulator/path_T=14_combined.pdf}
  \caption[labelInTOC]{Example Trajectories from \cref{eq:X_evolution_uo}
  using the parameter values from \cref{tab:regimes}. A)
  Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise C) Sub-Threshold-High-Noise D) Sub-Threshold-Low-Noise}
  \label{fig:regime_path_examples} 
\end{center} 
\end{figure}

@@

Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
(unknown) remaining time to spike, $\trem = \ts - t$. 
Starting from an arbitrary $t, x$, our remaining-cost
objective, $J[\a(\cdot); x,t]$, will be
\begin{equation}
J[\a(\cdot); x,t]  = 
\Exp\left[
\big(\trem - (\T-t) \big)^2 
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\, X_t = x
\right] 
\label{eq:OC_LS_cost_to_go}
\end{equation}
i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
Recall that in the closed-loop scenario, we assume that the value of $x = X_t$
is known to the controller.

\subsection{Hamilton-Jacobi-Bellman equation}
The Hamilton-Jacobi-Bellman (HJB) equation,
see \cite{Fleming1975,Evansb,Krylov2008}, \{ALEX: Should we also cite
Moehlis' papers that use/discuss HJB \cite{Nabi2013a,Danzl2009}???\}
associated with \cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$
% \footnote{Traditionally, the value function is denoted as $v$. We
% use $\v$ since we are working in a neural context, where $v$ is so often used
% to denote a voltage}
:
\begin{align}
\v(x,t) =&
 \min_{\a(\cdot)_{s \geq t}}
 \{J[\a(\cdot); x,t] \}
\notag
\\
=&
\min_{\a(\cdot)_{s \geq t}}
\Exp\left[
\big(\trem - (\T-t) \big)^2 
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\,X_t = x
\right]
\label{eq:OC_LS_mean_value}
\end{align}

Then $\v$ satisfies the following HJB partial differential equation (PDE) (see
\cite{Fleming1975,Evansb,Krylov2008} for details):
\begin{equation}
\begin{gathered}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2 \v(x,t) + \\
\min_{\a(x,t) \in [\amin,
\amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
\v(x,t) \Big\} = 0
\label{eq:OC_LS_mean_HJB}
\end{gathered}
\end{equation}
The special feature of \cref{eq:OC_LS_mean_HJB} in contrast to a generic
parabolic PDE is that it contains an embedded optimisation that depends on the
solution, $\v$. That is for each $x,t$ in the computational domain,
$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$. 
Here the we can solve for the optimal control, $\a^*(x,t)$ analytically as:
\begin{align}
\a^*(x,t)  =& \argmin_{\a \in [\amin, \amax]}\big\{\e \a^2 + (\m + \a-\tfrac{x}{\tc}) \di_x\v\big\}
\notag
\\
=&
\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
\label{eq:OC_LS_variance_energy_quadratic_control}
\end{align}
With this the HJB PDE becomes
\begin{equation}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v+
\e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
= 0
\label{eq:OC_LS_mean_HJB_bounded_control}
\end{equation}

Now, we need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$
then we have a spike now and we know that $\ts = t$. Thus
$$ \v(\xth,t) =
(t-\T)^2.$$ 
At the threshold, the value function equals the squared difference
between the desired spike time and the realized one.
Now for $x \ll .0$, we can assume that $\v$ is not
significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
justified if we impose artificial reflecting boundaries for the diffusion at $\xmin$. So we need to
determine for which $\xmin$ is a reflecting boundary consistent with the
dynamics. For example, we can take $\xmin$ to be two standard deviations
below the mean of the stationary distribution of the maximally
inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set $\xmin =
\tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. In fact, we further enforce that
$\xmin \leq -0.5$.
% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
% the BCs\ldots oh well.

Now we come to the Terminal Conditions (TCs) for $\v$. The idea is simple:
if we reach $\T$ without having spiked we apply maximum control in the
positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus: 
\begin{equation}\v(x,\T) =
\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:valuefun_TCs}
\end{equation} 
Note that we are making an approximation here
- that is we are ignoring the energy term, $\eps u^2$ in the objective for
$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
We discuss in more detail the validity of the approximation in
\cref{sec:effect_of_eps}.
Alternatively, we could impose this approximate terminal condition at some $t^+
> \T$.

We will see the quantity on the right hand side of \cref{eq:valuefun_TCs}
repeatedly so we will give it a special name.
\begin{equation}
\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:Trem_squared}
\end{equation} 
$\Ttwo$ is the second moment of the remaining time to reach the
threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
easily for all $x$ in the domain by solving a stationary backward Kolmogorov
equation. We show the details in \cref{sec:valuefun_TCs}.

@@

\subsection{Solutions of the HJB equation}
We set $\T=1.5$ and $\eps = .001$, so that the primary focus is the accurate
spiking and energy minimization considerations are secondary.

>>>>

\documentclass[12pt]{iopart}
%\documentclass{article}

@@

\usepackage{iopams}

% \newtheorem{thm}{Theorem}[section]
% \newtheorem{lemma}{Theorem}[thm]
% \theoremstyle{definition}
% \newtheorem{ex}{Example}[thm]
% \newtheorem{defn}{Definition}[thm]

@@

\title{Stochastic Optimal Control of Single Neuron Spike Trains}
\author{Alexandre Iolov$^1$, Susanne Ditlevsen$^2$, Andr\'e Longtin$^1$
		}	
\address{%
     (1) Department of Mathematics and Statistics, University of Ottawa, Ottawa,
    Canada
\\
    (2) Department of Mathematical Sciences, University of Copenhagen,
Copenhagen, Denmark
}
\ead{aiolo040@uottawa.ca}			
% \address{Department of Mathematics and Statistics, University of Ottawa, Ottawa, Canada 
% \\ Department of Mathematical Sciences, University of Copenhagen, Copenhagen,
% Denmark }
% \ead{$<$\href{mailto:aiolo040@uottawa.ca}
% 		{aiolo040 at uottawa dot ca}$>$}

% \date{\today}

\begin{abstract}

{\em Objective.} External control of exact spike timings in single neurons is
important for basic understanding of neuronal activity, and has potential in
brain-machine interfaces, for improving neural prostheses, and in medical
applications related to neuronal disorders, such as Parkinson's disease or
epilepsy. The goal of this paper is the design of optimal controlled electrical
stimulation patterns to a cell to achieve a target spike train under
physiological constraints not to damage tissue and minimize possible side
effects. {\em Approach.} We pose an optimal control problem to precisely specify
the spike timing in a noisy leaky integrate-and-fire model of a neuron. In
particular, we allow for the noise to be of arbitrary intensity. The optimal
control problem is solved using Dynamic Programming if the controller has access
to the voltage (closed-loop control) and using a Maximum Principle for the
transition density if the controller only has access to the spike timings (open
loop control). {\em Main results.} We have developed a stochastic optimal
control scheme to obtain precise spike timings, which is not limited by the
intensity of the noise in the system. Given this scope, the numerical demands
are non-trivial, but simulations show that the algorithms are potentially
feasible. It however remains to see if they can be successfully applied in a
real-time experimental setting. Furthermore, we show how wrong the control
strategy can perform if we assume that the noise is negligible. {\em
Significance.} More precise targeting and control of neural activity to achieve
optimal performance in a noisy environment, in which most neurons are embedded,
have great potential in health care. The main contribution is the online
feedback control of a noisy system through modulation of the input, taking into
account physiological constraints on the control, as well as incurred costs to
diminish adverse effects.

\end{abstract}
% \tableofcontents MSC KEYS: 49N90 90C39 93E20 93C20 93C95

@@

@@

Our objective, namely targeting exact spike times in single neurons, has been
considered mainly in the open-loop context, and in either absence of or for
small noise. In \cite{Ahmadian2011} they use the Spike Response Model,
\cite{Gerstner2002}, to control output target spike trains and implement their
scheme on pyramidal cells in mouse cortical slices. Their method is numerically
efficient and allows for the simultaneous control of many neurons. However the
efficiency of the numerics strongly relies on the assumption that the diffusion
noise, the value of $\b$ in \cref{eq:X_evolution_uo} below, is small enough, see
the discussion above and below equation 12 in \cite{Ahmadian2011}. They only
work with open-loop control. The difference between the objective in
\cite{Ahmadian2011} and what we consider is that they maximize the probability
of spiking at some given time $t^*$, whereas we minimize the mean squared
difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
Moehlis et al. \cite{Moehlis2006} work with a Phase Response Model to obtain
spikes at exact times, while keeping the root mean square of the input to a
minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
consider noise, though, and only work with open-loop control. In
\cite{Nabi2013}, the methods are implemented on brain slices of pyramidal
neurons of rat hippocampus. While the
Phase Response Curve Model is a parsimonious and effective way to describe a
neuron's response to a stimulus, it is only valid in the supra-threshold regime,
where the unstimulated neuron is periodically spiking. In \cite{Feng2003} they
use a leaky integrate-and-fire model and open-loop control. To
obtain a given spike time, they choose an objective of minimizing the
variance of the membrane potential at the desired spike time, while forcing the mean of
the membrane potential at this time point to equal the threshold. This provides
exact solutions since it does not involve first-passage times, but has the
drawback that there is a non-negligible probability that the obtained
spike time will be far from the desired spike time.

Our objective of imposing a certain timing sequence of the spike train using an
externally applied control is obtained in both the closed- and
open-loop setting, and we specifically include the noise in the
calculations of the controls. We restrict the controlled input to stay
within pre-specified bounds, and also include a cost function to
minimize intervention.
Our main contributions are that we specifically allow for a non-negligible
noise component in the neural activity when calculating the control, and
consider both open- and closed-loop controls. In particular we do not
restrict our attention to the autonomously-spiking, supra-threshold regime.

@@

\section{Problem Formulation}
A basic but useful model for the neural membrane potential evolution is the
noisy leaky-integrate-and-fire model:
\begin{equation}
\begin{gathered}
dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
\\
X(0) = 0,
\\
X(\ts) = \xth \implies
\begin{cases}
X(\ts^+) = 0 &
\end{cases}
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}
Here $X$ represents the membrane electric potential which decays to $0$
with a time constant of $\tc$, $dW$ is a Brownian motion increment scaled by
$\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
resets to $0$ and the process starts all over again. 
We write $\ts^+$ for the voltage's limit from the right at $\ts$.
Throughout the paper the threshold is set to one, $\xth = 1$.

Suppose that we have some control over the external current such that it can be
decomposed as
\begin{equation}
\Iext(t) = \m + \a(t)
\label{eq:current_mu_alpha}
\end{equation}
where $\m$ is an uncontrollable, but constant part of the external current and
$\a(t)$ is controllable, i.e.\ it can be chosen to achieve some goal.
A natural goal is to attempt to control the spike time, $\ts$.
That is how do we choose $\a(t)$ such that $\ts = \T$.
A natural optimal control objective to achieve this is the least squares solution
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
\label{eq:OC_LS_variance}
\end{equation}
where expectation is taken with respect to the distribution of the trajectories
of $X_t$.

Often the control has certain constraints. The most common are simple box
constraints:
\begin{equation}
\a(t) \in [\amin, \amax] \quad \forall t.
\label{eq:bound_constraints_alpha}
\end{equation}

% \subsection{Optimal Control Formulation}
In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
running cost based on the control. This regularizes the problem eliminating the
subtleties of singular-control situations and it serves to avoid excessive
control as well as to avoid excessive charge building up on the cell.

So we seek an optimal control, $\a^*$, that solves
\begin{align}
J[\a(\cdot)] =&
\Exp\left[
\e \int_0^\ts  \a^2(s) \intd{s}
+
(\ts - \T \big)^2 \right]
\label{eq:OC_LS_variance_energy}
\\ \notag
\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)].
\end{align}
where $\e$ measures how much weight we put on minimizing the energy cost.
If $\e = 0$, then we do not care at all about the expended cost.

It will often be the case that $\a$ means either a function or a
value of that function at a particular time. This function could be
random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
of $X$, it will be at least measurable with respect to the filtration of $X$,
see \cite{Oksendal2007} for the details. We will try to make clear below when we
are refering to $\a$ as a stochastic process and when we are merely refering to its particular
value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
refers to the function, while $\a(s)$ refers to that function's value, possibly
random, at time $s$.

We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
control. In closed-loop control, the value of $X_t$ at time $t$ is observable
and can be used in determining the control. In open-loop control only the spike
times are observable. In the closed-loop context, we write the control as $$\a =
\a(x,t)$$ to indicate its dependence on $X_t$ and to express that it will be
updated based on the time-course of $X$. In the open-loop context, we write $$\a
= \a(t)$$ to indicate that $\a(\cdot)$ is decided for all times at time 0. The
techniques used to obtain the optimal controls in the two scenarios will be
different. For the closed-loop scenario we use Dynamic Programing,
see \cite{Fleming1975}, while for the open-loop scenario we use a form of the
Maximum Principle applied on the transition density of the controlled process,
see \cite{Borzi2012}.

Crucially, we assume that the model parameters, $\m,\tc,\b$ in
\cref{eq:X_evolution_uo} are known. This is a strong assumption and will be
discussed later.

\subsection{Parameter Regimes}
Different  parameter regimes can be envisioned given \cref{eq:X_evolution_uo},
depending on whether the noise intensity, $\b$, is
relatively high or low, and whether the external, uncontrollable bias current,
$\m$, induces spikes in the absence of noise or not. 
Spikes will occur in the absence of noise, if and only
if $ \m > \frac 1\tc,$ which is called the supra-threshold regime. When $\m
\leq 1/\tc$, the regime is called the sub-threshold regime.
In addition we will impose two values of $\b$, which we
will call high-noise and low-noise, respectively.
Example values for each parameter regime are given in \cref{tab:regimes}, and we
visualize a single path for each regime in \cref{fig:regime_path_examples}.
\begin{table}
\begin{tabular}{|l||{c}|{c}|}
\hline
\backslashbox{$\m$}{$\b$}
& $1.5$ & $0.3$ \\
\hline
$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\
\hline
$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
\hline
\end{tabular}
\caption{Regime labels and example values. Note that for the numerical
experiments below, we use $\tc = 0.5$}
\label{tab:regimes}
\end{table}
%\usepackage{graphics} is needed for \includegraphics
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=0.99\textwidth]{Figs/PathSimulator/path_T=14_combined.pdf}
  \caption[labelInTOC]{Example Trajectories from \cref{eq:X_evolution_uo}
  using the parameter values from \cref{tab:regimes}. A)
  Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise C) Sub-Threshold-High-Noise D) Sub-Threshold-Low-Noise}
  \label{fig:regime_path_examples}
\end{center}
\end{figure}

@@

Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
(unknown) remaining time to spike, $\trem = \ts - t$. Starting from an arbitrary
$t, x$, our remaining-cost objective, $J[\a(\cdot); x,t]$, will be
\begin{equation}
J[\a(\cdot); x,t]  =
\Exp\left[
\big(\trem - (\T-t) \big)^2
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\, X_t = x
\right]
\label{eq:OC_LS_cost_to_go}
\end{equation}
i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
Recall that in the closed-loop scenario, we assume that the value of $X_t$
is known to the controller. A similar problem is discussed analytically at
length in the book on optimal control by Peter Whittle, \cite{Whittle1996},
although there is no discussion there of the numerics required to solve it.

\subsection{Hamilton-Jacobi-Bellman equation}
The Hamilton-Jacobi-Bellman (HJB) equation,
see \cite{Fleming1975,Whittle1996} or \cite{Danzl2009,Nabi2013a} for a neuroscience
context, associated with
\cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$:
\begin{align}
\v(x,t) =&
 \min_{\a(\cdot)_{s \geq t}}
 \{J[\a(\cdot); x,t] \}
\notag
\\
=&
\min_{\a(\cdot)_{s \geq t}}
\Exp\left[
\big(\trem - (\T-t) \big)^2
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\,X_t = x
\right]
\label{eq:OC_LS_mean_value}
\end{align}

Then $\v$ satisfies the following HJB partial differential equation (PDE):
\begin{equation}
\begin{gathered}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2 \v(x,t) + \\
\min_{\a(x,t) \in [\amin,
\amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
\v(x,t) \Big\} = 0
\label{eq:OC_LS_mean_HJB}
\end{gathered}
\end{equation}
The special feature of \cref{eq:OC_LS_mean_HJB} in contrast to a generic
parabolic PDE is that it contains an embedded optimisation that depends on the
solution, $\v$. For each $x,t$ in the computational domain,
$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$.
Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
\begin{align}
\a^*(x,t)  =& \argmin_{\a \in [\amin, \amax]}\big\{\e \a^2 + (\m + \a-\tfrac{x}{\tc}) \di_x\v\big\}
\notag
\\
=&
\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
\label{eq:OC_LS_variance_energy_quadratic_control}
\end{align}
With this the HJB PDE becomes
\begin{equation}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v+
\e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
= 0
\label{eq:OC_LS_mean_HJB_bounded_control}
\end{equation}

We need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$ then we
have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
threshold, the value function equals the squared difference between the desired
spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
justified if we impose artificial reflecting boundaries for the diffusion at
$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
consistent with the dynamics. For example, we can take $\xmin$ to be two
standard deviations below the mean of the stationary distribution of the
maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
that $\xmin \leq -0.5$.
% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
% the BCs\ldots oh well.

To determine the Terminal Conditions (TCs) for $\v$, the idea is
simple: if we reach $\T$ without having spiked we apply maximum control in the
positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus:
\begin{equation}\v(x,\T) =
\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:valuefun_TCs}
\end{equation}
Note that we are making an approximation here
- we are ignoring the energy term, $\eps u^2$, in the objective for
$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
We discuss in more detail the validity of the approximation in
\cref{sec:effect_of_eps}.
Alternatively, we could impose this approximate terminal condition at some $t^+
> \T$.

We will see the quantity on the right hand side of \cref{eq:valuefun_TCs}
repeatedly so we will give it a special name.
\begin{equation}
\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:Trem_squared}
\end{equation}
$\Ttwo$ is the second moment of the remaining time to reach the
threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
easily for all $x$ in the domain by solving a stationary backward Kolmogorov
equation. We show the details in \ref{sec:valuefun_TCs}.

@@

\subsection{Solutions of the HJB equation}
We set $\T=1.5$ and $\eps = 0.001$, so that the primary focus is the accurate
spiking and energy minimization considerations are secondary.
